A Bayesian ranking statistic to find high-mass black holes in LIGO--Virgo data

### 2021-04-08 main.tex
Updates from Overleaf

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{Introduction}
Since the 1970s, there has been a steady accumulation of evidence for stellar mass ($\text{M}_\text{BH} < 10^{2} \ \msun$) and supermassive black holes ($\text{M}_\text{BH} > 10^{5} \ \msun$)~cite{Webster:1972:Natur, Balick:1974:ApJ, Ghez:1998:ApJ, Genzel:2010:RvMP, Abbott:2019:PhRvX, EventHorizonTelescopeCollaboration:2019:ApJL, Abbott:2020:arXiv}. However, there is a deficiency of observational evidence for black holes in the intermediate-mass range $10^{2} - 10^{5}\ \text{M}_{\odot}$. Given the existence of supermassive black holes detected in the early universe, it is likely for intermediate-mass black holes to have also existed~\cite{Banados:2018:Natur}. However, we only have a handful of promising candidates~\cite{Greene:2004:ApJ, Graham:2013:ApJ, Mezcua:2017:IJMPD, Koliopanos:2017:mbhe, Lin:2020:ApJL, Greene:2020:ARA&A}, and one confirmed intermediate-mass black hole, the remnant of the gravitational wave event GW190521 with a mass of $142^{+28}_{-16}\ \msun$~\cite{Abbott:2020:PhRvL}. The concrete discovery of more intermediate-mass black holes will bridge the observational gap and illuminate our understanding of galaxy and supermassive black hole formation.

The two main proposed formation channels for Intermediate-mass black holes are the light-seed and the heavy-seed channels. The light-seed channel involves dynamic interactions of low-mass seeds, e.g. hierarchical mergers of stellar-mass black holes~\cite{}, runaway collisions of stars~\cite{}.  The heavy-seed channel involves the direct collapse of high-mass seeds, e.g. the direct collapse of giant metal-free Population III stars~\cite{}, early protogalaxies' collapse~\cite{}.  Although promising pathways, there are challenging obstacles both scenarios must overcome to pathways that can form intermediate-mass black holes.

Hierarchical mergers in active galactic nuclei migration-traps, globular and stellar clusters may be a feasible light-seed method for black holes to accumulate mass and reach the intermediate-mass range. \rs{\sout{However, when black holes merge, they can receive recoil-kicks up to $\sim$2,500 km/s~\cite{}. Hence, the merger remnants require their recoil kick to be less than its environment's escape velocity to undergo successive mergers. This tight constraint makes hierarchical mergers a challenging scenario.}}

One proposed heave-seed pathway involves the direct collapse of massive, metal-free Population III stars. Typically, massive stars $> 100 \ \msun$ can eject more than half of their mass in the form of radiatively driven stellar winds~\cite{}. However, at zero metallicity, mass loss through stellar winds is negligible \cite{Kudritzki 2000}. Consequently,  Population III stars have the opportunity to complete their life-cycles without losing a significant portion of their mass. Moreover, Population III stars with masses above $230 \ \msun$ will avoid electron-positron pair-instability. Therefore, oxygen and silicon burning will not drive an explosion, allowing the star to undergo a complete collapse to leave behind an intermediate-mass black hole remnant with a mass of $120-280 \ \msun$, dependent on the initial star mass \cite{Bond et al. 1984}. As the fraction of Population III stars with masses above $230 \ \msun$ is uncertain, the number of expected intermediate-mass black holes from these stars is also ambiguous~\cite{Coleman Miller, Formation Scenarios for Intermediate-Mass Black Holes}.

To verify and determine details on which formation channel is favoured, a catalog of observed intermediate-black holes and their masses is required. However, measuring the mass of black holes can be challenging.

\rs{\sout{It is possible to measure the mass of nearby supermassive black holes by directly observing the dynamic motion of stars nearby the supermassive black hole (such as Sgr A$^*$~). However, this dynamic measurement method is challenging to utilise for  intermediate-mass black holes as their sphere of influence is too small to resolve motions~\cite{}. Additionally, dynamic-measurements that have alluded to  intermediate-mass candidates, e.g. the candidate at the center of the globular cluster 47 Tucanae~\cite{kiziltan 2017}, can also be attributed to simpler models without a central  intermediate-mass black hole~\cite{freire 2017}. }}

\rs{\sout{In a similar vein, although the M-$\sigma$ and M-L relations (relations between the spheroidal bulge of a galaxy's velocity dispersion $\sigma$,  luminosity L, and its central black hole's mass M) can estimate the mass of supermassive black holes~\cite{Ferrarese & Merritt 2000 and Gebhardt et al. 2000}, they cannot accurately estimate the intermediate-masses. The M-$\sigma$ and M-L relations are not calibrated for low-mass galaxies that might host an central intermediate-mass black hole~\cite{J. E. Greene and L. C. Ho The Mass Function of Active Black Holes in the Local Universe, E. C. Moran, K. Shahinyan, H. R. Sugarman, D. O. Vélez and M. Eracleous Black Holes At the Centers of Nearby Dwarf Galaxies}. Table 2 of~\citet{mercuza} displays a list of intermediate-mass candidates found with these relations. }}

\rs{\sout{Another method to estimate large black hole masses is with reverberation mapping, which estimates black hole virial masses by studying their nearby gasses' velocities~\cite{reverberation mapping}. Reverberation mapping has shown to estimate both supermassive black hole and intermediate-mass black hole virial masses successfully. However, this method fails to measure the systematic uncertainties on the lower-end of the intermediate-mass spectrum~\cite{}. Additionally, cloud-variability and the possibility of anisotropic emission deride the credibility of intermediate-mass candidates found with reverberation mapping~\cite{}. }}

The most promising intermediate-mass candidates found with electromagnetic-spectra are Ultra-Luminous X-ray sources with luminosities $> 10^{39} \text{erg s}^{-1}$ (e.g. HLX-X1)~\cite{}. \rs{\sout{While stellar mass black holes are too small to generate such luminosities with Eddington accretion, intermediate mass black holes can accrete at rates large enough to account for the high luminosities~\cite{}. However, some argue that some stellar mass black holes may exceed the Eddington limit~\cite{}, and others argue that the X-ray emission might be from a collimated jet leading to an inaccurate mass calculation~\cite{}.}}

\Finally, \rs{\sout{another channel to measure black hole masses is by identifying}} Gravitational waves from compact binaries coalesces (CBCs) \rs{are clean channels to measure BH masses} \rs{\sout{ inferring their source parameters~\cite{}.}} Mergers involving either a binary system with at least one intermediate-mass component, or a system whose merger results in an intermediate-mass remnant will have large initial total masses $\text{M}_{T}$. As a binary's $\text{M}_{T}$ is associated with its merger frequency, $f\sim \text{M}_T^-1$, systems with large $\text{M}_{T}$ have very low merger frequencies $f < 100\ \text{Hz}$. Hence, ground based gravitational wave detectors ($\sim 10 - 1000\ \text{Hz}$) are sensitive to the last milliseconds of merging systems with $100 \msun < \text{M}_{T} < 500 \msun$, while space-based detectors ($\sim 0.1 - 10\ \text{Hz}$) can study the full signals of merging systems with $10^4 \msun < \text{M}_{T} < 10^7 \msun$.

@@

After calculating the \bcr for a set of background triggers and simulated triggers from as stretch of detector-data (a data chunk), we can compute probability distributions for the background and simulated triggers, $p_b(\bcr)$ and $p_s(\bcr)$. We expect the background trigger and simulated signal \bcr values to favour the incoherent glitch and the coherent signal hypothesis, respectively. Ideally, these distributions representing two unique populations should be distinctly separate and have no overlap in their \bcr values. The prior odds parameters $\alpha$ and $\beta$ from Eq.~\ref{eq:bcr} help separate the two distributions. Altering $\alpha$ translates the \bcr probability distributions while adjusting $\beta$ spreads the distributions. Although Bayesian hyper-parameter estimation can determine the optimal values for $\alpha$ and $\beta$, an easier approach is to adjust the parameters for each data chunk's \bcr distribution. In this study, we tune $\alpha$ and $\beta$ to maximally separate the \bcr distributions for the background and simulated triggers.

@@

\section{Marginalizing over PSD statistical uncertainties}\label{sec:psd-marginalization}
To generate the results in Fig.~\ref{fig:bcrCdf}, we applied a post-processing step to marginalize the uncertainty in the PSD. In Fig.~\ref{fig:bcrCdfUnmarginalized}, we show the results if this post-processing step is not applied. Clearly, marginalizing over uncertainty in the PSD yields an improvement in the separation of the noise and signal distributions. Quantitatively, this results in a reduction in the rate of false alarms from X to Y and a reduction in the rate of false positives from X to Y in the August 13 - 21, 2017 time-frame of data. For the entirety of O2 PSD marginalisation resulted in a reduction of X, or an $x\%$ improvement in search efficency.

>>>>

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{Introduction}
Since the 1970s, there has been a steady accumulation of evidence for stellar mass ($\text{M}_\text{BH} < 10^{2} \ \msun$) and supermassive black holes ($\text{M}_\text{BH} > 10^{5} \ \msun$)~\cite{Webster:1972:Natur, Balick:1974:ApJ, Ghez:1998:ApJ, Genzel:2010:RvMP, Abbott:2019:PhRvX, EventHorizonTelescopeCollaboration:2019:ApJL, Abbott:2020:arXiv}. However, there is a deficiency of observational evidence for black holes in the intermediate-mass range $10^{2} - 10^{5}\ \text{M}_{\odot}$. Astronomers have detected a handful of intermediate-masses candidates using reverberation mapping, taking direct kinematic measurements, applying  M-$\sigma$ and M-L relations, and also studying  X-ray luminosity and spectra~\cite{Greene:2004:ApJ, Graham:2013:ApJ, Mezcua:2017:IJMPD, Koliopanos:2017:mbhe, Lin:2020:ApJL, Greene:2020:ARA&A}. However, these observations are highly uncertain and can be attributed to sources without intermediate-mass black holes~\cite{kiziltan2017, freire2017}. In the last few years, astronomers have found definitive evidence for two intermediate-mass black holes, (i) the $142^{+28}_{-16}\ \msun$ remnant observed from the gravitational wave event GW190521~\cite{Abbott:2020:PhRvL}, and the $5.5^{+1.7}_{-0.9}\times10^{4}\ \msun$ black hole identified from gravitational lensing in the light curve of GRB950830~\cite{paynter_evidence_2021}. The concrete discovery of more intermediate-mass black holes will bridge the observational gap and illuminate our understanding of galaxy and supermassive black hole formation. Additionally, a catalog of observed intermediate-mass black holes can act as probes to their formation environments, such as in the accretion disks of active galactic nuclei~\cite{}, the centers of dense stellar clusters~\cite{}, and even Population-III stars~\cite{}.

% Given the existence of supermassive black holes detected in the early universe, it is likely for intermediate-mass black holes to have also existed~\cite{Banados:2018:Natur}. However, we only have a handful of promising candidates~\cite{Greene:2004:ApJ, Graham:2013:ApJ, Mezcua:2017:IJMPD, Koliopanos:2017:mbhe, Lin:2020:ApJL, Greene:2020:ARA&A}, and two confirmed intermediate-mass black holes, (i) the $142^{+28}_{-16}\ \msun$ remnant observed from the gravitational wave event GW190521~\cite{Abbott:2020:PhRvL}, and the $5.5^{+1.7}_{-0.9}\times10^{4}\ \msun$ black hole identified from gravitational lensing in the light curve of GRB950830~\cite{paynter_evidence_2021}. The concrete discovery of more intermediate-mass black holes will bridge the observational gap and illuminate our understanding of galaxy and supermassive black hole formation. Additionally, a catalog of observed intermediate-mass black holes can act as probes to their formation environments, such as in the accretion disks of active galactic nuclei~\cite{}, the centers of dense stellar clusters~\cite{}, and even Population-III stars~\cite{}.

% However, measuring the mass of black holes can be challenging. Several observational techniques to identify intermediate mass black holes

% It is possible to measure the mass of nearby supermassive black holes by directly observing the dynamic motion of stars nearby the supermassive black hole (such as Sgr A$^*$~). However, this dynamic measurement method is challenging to utilize for  intermediate-mass black holes as their sphere of influence is too small to resolve motions~\cite{}. Additionally, dynamic-measurements that have alluded to  intermediate-mass candidates, e.g. the candidate at the center of the globular cluster 47 Tucanae~\cite{kiziltan2017}, can also be attributed to simpler models without a central intermediate-mass black hole~\cite{freire 2017}. }}

% \rs{\sout{In a similar vein, although the M-$\sigma$ and M-L relations (relations between the spheroidal bulge of a galaxy's velocity dispersion $\sigma$,  luminosity L, and its central black hole's mass M) can estimate the mass of supermassive black holes~\cite{Ferrarese & Merritt 2000 and Gebhardt et al. 2000}, they cannot accurately estimate the intermediate-masses. The M-$\sigma$ and M-L relations are not calibrated for low-mass galaxies that might host an central intermediate-mass black hole~\cite{J. E. Greene and L. C. Ho The Mass Function of Active Black Holes in the Local Universe, E. C. Moran, K. Shahinyan, H. R. Sugarman, D. O. Vélez and M. Eracleous Black Holes At the Centers of Nearby Dwarf Galaxies}. Table 2 of~\citet{mercuza} displays a list of intermediate-mass candidates found with these relations. }}

% \rs{\sout{Another method to estimate large black hole masses is with reverberation mapping, which estimates black hole virial masses by studying their nearby gasses' velocities~\cite{reverberation mapping}. Reverberation mapping has shown to estimate both supermassive black hole and intermediate-mass black hole virial masses successfully. However, this method fails to measure the systematic uncertainties on the lower-end of the intermediate-mass spectrum~\cite{}. Additionally, cloud-variability and the possibility of anisotropic emission deride the credibility of intermediate-mass candidates found with reverberation mapping~\cite{}. }}

% The most promising intermediate-mass candidates found with electromagnetic-spectra are Ultra-Luminous X-ray sources with luminosities $> 10^{39} \text{erg s}^{-1}$ (e.g. HLX-X1)~\cite{}. \rs{\sout{While stellar mass black holes are too small to generate such luminosities with Eddington accretion, intermediate mass black holes can accrete at rates large enough to account for the high luminosities~\cite{}. However, some argue that some stellar mass black holes may exceed the Eddington limit~\cite{}, and others argue that the X-ray emission might be from a collimated jet leading to an inaccurate mass calculation~\cite{}.}}

Gravitational waves from compact binaries coalesces (CBCs) are clean channels to measure BH masses. Mergers involving either a binary system with at least one intermediate-mass component, or a system whose merger results in an intermediate-mass remnant will have large initial total masses $\text{M}_{T}$. As a binary's $\text{M}_{T}$ is associated with its merger frequency, $f\sim \text{M}_{T}^{-1}$, systems with large $\text{M}_{T}$ have very low merger frequencies $f < 100\ \text{Hz}$. Hence, ground based gravitational wave detectors ($\sim 10 - 1000\ \text{Hz}$) are sensitive to the last milliseconds of merging systems with $100 \msun < \text{M}_{T} < 400 \msun$~\cite{}, while space-based detectors ($\sim 0.1 - 10\ \text{Hz}$) can study the full signals of merging systems with $10^4 \msun < \text{M}_{T} < 10^7 \msun$~\cite{}.

@@

After calculating the \bcr for a set of background triggers and simulated triggers from as stretch of detector-data (a data chunk), we can compute probability distributions for the background and simulated triggers, $p_b(\bcr)$ and $p_s(\bcr)$. We expect the background trigger and simulated signal \bcr values to favor the incoherent glitch and the coherent signal hypothesis, respectively. Ideally, these distributions representing two unique populations should be distinctly separate and have no overlap in their \bcr values. The prior odds parameters $\alpha$ and $\beta$ from Eq.~\ref{eq:bcr} help separate the two distributions. Altering $\alpha$ translates the \bcr probability distributions while adjusting $\beta$ spreads the distributions. Although Bayesian hyper-parameter estimation can determine the optimal values for $\alpha$ and $\beta$, an easier approach is to adjust the parameters for each data chunk's \bcr distribution. In this study, we tune $\alpha$ and $\beta$ to maximally separate the \bcr distributions for the background and simulated triggers.

@@

\section{Marginalizing over PSD statistical uncertainties}\label{sec:psd-marginalization}
To generate the results in Fig.~\ref{fig:bcrCdf}, we applied a post-processing step to marginalize the uncertainty in the PSD. In Fig.~\ref{fig:bcrCdfUnmarginalized}, we show the results if this post-processing step is not applied. Clearly, marginalizing over uncertainty in the PSD yields an improvement in the separation of the noise and signal distributions. Quantitatively, at a threshold $\bcr_T=0$ the post-processing step results in a reduction in the number of background $\bcr > \bcr_T$ from $60.7\%$ to $25.28\%$ in the August 13 - 21, 2017 time-frame of data. For the entirety of O2 PSD marginalization resulted in a $49.26\%$ improvement in search efficiency.
